Towards practical dynamic induction control of wind farms: analysis of optimally controlled wind-farm boundary layers and sinusoidal induction control of first-row turbines

### 2018-04-17 revision/reply_to_reviewers.tex

\documentclass[]{article}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

%\maketitle

\section*{Response to Reviewer 1}

We thank the reviewer for his/her constructive comments and are pleased with the positive assessment or our work. We have addressed the specific comments made by the reviewer as described below. We hope that the revised manuscript can now be accepted for publication.

\dotfill

\textbf{1. Reviewer:} \textit{One comparison that came to mind reading the paper, is that the results show very little value in optimizing the most-downstream turbines, and in general, improvement in power comes from modifications upstream, and that self-optimization is not possible. This stands in contrast to a result such as:}

\textit{Ciri, Umberto, Mario Rotea, Christian Santoni, and Stefano Leonardi. “Large Eddy Simulation for an Array of Turbines with Extremum Seeking Control.” In American Control Conference. Boston, MA, 2016. }

\textit{Where the TSR of downstream turbines is re-optimized for wake conditions (and the upstream turbine is left as is at the end of the optimization). It would seem the difference in modeling methods and/or how turbine control is implemented yields the different results, but I believe it would be worth discussing the difference, for example around the paragraph beginning with "The figure shows that the first row (R1)..." on page 12.}

\textbf{Response:} The reason for this different behavior of the downstream vs upstream turbines is twofold:
\begin{enumerate}
	\item The Extremum Seeking Control (ESC) approach by Ciri \emph{et al.} differs fundamentally from our approach. In our approach, we simultaneously optimize all control variables over a given time horizon, which allows for full dynamic cooperation and interaction between all turbines. In the ESC, controls are optimized one by one in the sense that ESC is initiated on the first row turbine and will be activated on the subsequent row only after its upstream neighbor has converged to a steady state solution. In this way, dynamic control and interaction between controls in different turbines are discarded a priori. Furthermore, although this is not an inherent limitation of the ESC approach, it seems that Ciri \emph{et al.} optimize for individual turbine power instead of total wind-farm power (see Eq. 10 $P_i$ is used instead of $\sum_i P_i$). Note that in Santoni \emph{et al.} (from the same research group as Ciri \emph{et al.}), where a Jensen model is used to optimize total wind-farm power using the individual tip speed ratios, it is also found that after full-farm optimization the downstream turbine operates at its individually optimal TSR (consistent with our observations).

	\item In our modeling approach, we make abstraction of detailed turbine dynamics in the sense that we directly control the thrust coefficient $C_T'$ instead of the generator torque or tip speed ratio. In this sense, the optimal thrust coefficient value of a turbine with no downstream neighbors can be analytically found to be $C_T' = 2$. In Ciri \emph{et al.}, the authors indicate that the reason why the TSR in the downstream rows (around 6.5) does not correspond to the individual optimal TSR (7.5) of individual turbines is that the TSR depends on a reference velocity and (direct quotation) ``As a consequence, the TSR is not a meaningful indicated of a waked turbine operating condition".

\end{enumerate}

Summarizing, the inverse behavior of the control dynamics compared to Ciri \emph{et al.} can be attributed to the difference in approach to wind-farm control (dynamic full-farm optimization vs steady one-by-one optimization) as well as the implementation of the turbine control ($C_T'$ vs $\lambda$). We believe that, if we were to constrain ourselves to steady control and would control a similarly defined $\lambda$ as in Ciri \emph{et al.}, our observations would be similar to the latter study. Note that Ciri \emph{et al.} do not increase wind-farm power over the case with turbines operating at locally optimal TSR, whereas our dynamic approach allows to improve power significantly over the case with locally optimal operation.

In our opinion, the current study (full-farm dynamic wind-farm optimization) and that of Ciri \emph{et al.} (providing a practical ESC method for optimizing individual turbines) cannot be directly compared. However, we understand the contrast in the results and therefore added a small comment to the revised manuscript (WM - WEET NIET OF WE DIT WEL ECHT MOETEN DOEN, COMMENTS KOMT WAT VERLOREN UIT IN MANUSCRIPT, MISSCHIEN GEWOON REVIEWER OVERTUIGEN DAT WE DIT NIET INCLUDEN?) (l. ??, p. ??):

``
... is not monotonous. Note that these observations are in contrast to the study of Ciri \emph{et al.} (REF), where Extremum Seeking Control is applied to optimize an array of three turbines and it was found that the set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are re-optimized for wake conditions. This contrast can mainly be attributed to our dynamic control approach, which allows for proper improvement of downstream flow conditions (which is not the case for set-point control as in Ciri \emph{et al.}), and that our ADM makes abstraction of any re-optimization of TSR.
''

WM - KAN WEL GOED ZIJN DAT ROTEA HIER DE REVIEWER IS - ZIEN DAT WE NIETS FOUT ZEGGEN. ANTWOORD IS NOG NIET ZO SCHERP ALS IK HET ZELF WIL. VOORAL IN ENUMERATE NOG WAT TE KNIPPEN

\dotfill

\textbf{2. Reviewer:} \textit{Figure 3/related text: Would another way to describe Ct2 vs Ct3 be that Ct2 can only lower the thrust, while Ct3 is allowed to raise it?}

\textbf{Response:} Indeed, this is correct. We have mentioned this explicitly in the revised manuscript as follows (p. ??, l. ??):

``
... and the maximal thrust coefficient $C_{T,\rm max}' = 2$ or 3, \textbf{with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. 5). 
''

\dotfill

\textbf{3. Reviewer:} \textit{``... NREL 5MW rotor with a 50\% increase in chord length ...'' does this imply the method is currently assuming the chord length is variable? Could this not be achieved by a change in pitch angle?}

\textbf{Response:} No, the method does not imply a variable chord length.

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. ??, l. ??):
``
be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A)
''

\clearpage
\section*{Response to Reviewer 2}

\end{document}

### 2018-04-17 revision/reply_to_reviewers.tex

\textbf{Response:} No, the method does not imply a variable chord length.

Current turbines are designed to approach maximum $C_T'$ values around 2, corresponding to the Betz limit. Although this maximum value can be increased by increasing the operational tip speed ratio (TSR), aiming to do so by adapting the pitch angle would inevitably lead to severe efficiency losses due to stall on the turbine blades. Therefore, we provide an example of how an alternative turbine design (i.e. with an increased chord length and operational TSR) could attain a maximum $C_T'$ of 3.5. Given such a turbine design, achieving thrust ratings $C_T' < 3.5$ is straightforward by pitching blades towards the feather position.

We have slightly modified the statement in the revised manuscript to avoid any confusion with regard to possibly having the chord length as a control variable. (p. ??, l. ??):
``
be \textbf{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in blade chord length and an operational tip speed ratio 25\% higher than 
the original design value (see Goit and Meyers, 2015, Appendix A)
''

\clearpage
\section*{Response to Reviewer 2}
We thank reviewer 2 for reading our work and for providing detailed feedback which has definitely improved the quality of the manuscript. We are pleased with his/her positive assessment of our research and have addressed the specific comments in the manuscript as described below. We hope that the revised manuscript can now be accepted for publication.

\textbf{Reviewer: } \textit{...LES setup is described/illustrated properly - except of the characteristics of the turbines/actuator disks considered. Would be nice to have the size of the disks explicitly noted (as they are somewhat hidden in Figure 2) to have a more clear scale of the considered wind farm.}

\textbf{Response: } Indeed, we seem to have overlooked specifying the exact turbine dimensions in Section 2.2. We have updated the manuscript as follows:

``
... 12 rows by 6 columns. \textbf{The wind turbines have a hub height $z_h = 100$~m with a rotor diameter $D = 100$~m, and are spaced 6$D$ apart in both axial and transversal directions.}
''

\textbf{Reviewer: } \textit{While describing the case setup in Section 2.2, the "flow advancement time", $T_A$ (also
	referred in Figure 1) is considered as half of the prediction horizon $T$. Would $T_A$ (and
	therefore $T$) be inflow dependent as the time delay (the time it takes for particles to
	move from the upstream to downstream turbine(s))? Have you investigated if changing
	$T$ (and/or $T_A$) has any effects on the resulting optimum $C_T$ set-points and on the power
	gain?}

\textbf{Response: }

\textbf{Reviewer: } \textit{As clearly seen in Figure 4c and 4d, there is a significant increase in turbulence
	further downstream. In addition to the TKE and the transport, would be nice to have the
	turbulence intensity TI values (as listed later on page 20, 10% for the baseline case),
	both for the baseline case and the maximum added TI reported - possibly somewhere
	around Figure 4. That again would give an indication on the applicability compared to
	the field values observed. Also note the typo in the caption of Figure 4: after c) all the
	subplots are marked to be continuously c).}

\textbf{Response: }

\textbf{Reviewer: } \textit{On page 10, around line 10, the argument of "upstream actions do not require a
	specific downstream response in order to increase power in that downstream row",
	which is also paraphrased in the conclusions, needs to be elaborated. This rather
	broad conclusion seem to oversee the probability of the curtailment of the downstream
	turbine where down-regulation might be inevitable for certain CT set-points assigned
	to downstream turbine(s) in the resulting optimization. Could be partially true for the
	investigated C3t5 case since there observed very limited curtailment even at the most
	upstream turbine (as in Figure 3b). However, also seen in Figure 8b (except of the
	very last row as the authors indicated), there seem to be still a difference between on
	the power gain at turbine R11 for the scenarios of R1-R10 and R1-R11. Narrowing the argument to the considered case or very little to no downstream curtailment CT
	distributions is suggested.}

\textbf{Response: }

\textbf{Reviewer: } \textit{On page 13, line 13, "the presence of the flow invariant features of the control signals"
	needs further justification as Figure 11 would also depend on how variant the flow
	features are in the simulations. That should include both the spatial and temporal
	variance within the 30-min window. As far as the field measurements are concerned,
	high spatial and temporal correlations are observed. For the former, Figure 2 gives a
	brief idea about the wind speed range between the columns, that can be referred here.
	For the latter, time series or relevant temporal statistics can be presented to assess the
	randomness and strengthen the hypothesis.}

\textbf{Response: }

\textbf{Reviewer: } \textit{On page 17, around line 5, a very nice example on how to implement the optimized
	sinusoidal CT is presented. The practical examples can be further improved by a
	short discussion on the expected response time of such increases in tip speed ratio
	on a machine with high inertia. That would put the estimated sine wave period into
	perspective as well.}

\textbf{Response: }

\textbf{Reviewer: } \textit{For Section 4.2.3, the header "Full-scale wind farm test" is a bit misleading... Suggest
	to change to "Full-scale wind farm simulations (in LES)" instead.}

\textbf{Response: }

\textbf{Reviewer: } \textit{On Figure 19, why would the power decrease after Row 5 for the sinusoidal case?}

\textbf{Response: }

\textbf{Reviewer: } \textit{Page 22 around line 5, the (inevitable) discussions on loads are included. In addition
	to the loads on the controlled upstream turbine, Figure 20(b) indicates partial wakes
	on the further downstream rows of turbines. Therefore, the section should be improved
	by highlighting the probable increase in fatigue loading for not just turbine(s) R1 but for
	the downstream rows as well, possibly starting as early as R3.}

\textbf{Response: }

\textbf{Reviewer: } \textit{On the grammatical note, the manuscript is clear and easy to follow. The only comment
	might be on the use of Sect. or Section; Fig. or Figure references.}

\textbf{Response: }

### 2018-05-17 revision/revised_WM01.tex
First draft of replies

@@

\subsection{Case setup}
The wind farm considered in MM17 has an aligned pattern of 12 rows by 6 columns, with 6 rotor diameters spacing in axial and tranversal directions. The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points. A snapshot of the streamwise velocity field is shown in Fig. \ref{fig:flowfield_MM17}. The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240$ s and a flow advancement time of $T_A = T/2 = 120$ s, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.

@@

A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case. Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq. \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 (underinductive or overinductive, see Eq. \ref{eq:boxct_constraint}). Cases are denoted as C<X>t<Y>, where X and Y represent $C_{T,\text{max}}'$ and $\tau$ respectively, e.g. C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s.

@@

Figures \ref{fig:flowfield_statistics}c shows sideviews of turbulence kinetic energy. The figure shows an increase in turbulence throughout the entire wind farm, spreading to the internal boundary layer above the turbines. Note specifically the sharp increase in turbulence in the core wake region behind the first-row turbine, for which an enhanced recovery was found as discussed above.

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figure4}
	\caption{Time-averaged flow field quantities of simulation results from MM17. \emph{Left (1):} Time averages of reference simulations with $C_T' = 2$ for all turbines. \emph{Right (2):} Difference $\Delta$ between reference and optimal control (C3t5) simulation, defined as $\Delta = X_{\rm C3t5} - X_{\rm REF}$ for any variable $X$. \emph{a):} Axial velocity $U_x$ (planview at hub height). \emph{b):} Axial velocity $U_x$ (sideview through turbine column). \emph{c):} Turbulence kinetic energy $k = (1/2)(\overline{\widetilde{u}'_x \widetilde{u}'_x} + \overline{\widetilde{u}'_y \widetilde{u}'_y} + \overline{\widetilde{u}'_z \widetilde{u}'_z})$ (sideview). \emph{c):} Turbulence top-down transport $ - \overline{\widetilde{u}'_x\widetilde{u}'_z}$. \emph{c):} Mean-flow top-down transport $ - \widetilde{U}_x \widetilde{U}_z$ (sideview). \emph{c):} Turbulence horizontal transport $\overline{\widetilde{u}'_x \widetilde{u}'_y}$ (planview). \emph{c):} Mean-flow horizontal transport $\widetilde{U}_x \widetilde{U}_y$ (planview). Black lines indicate wind turbine locations. Simulation results are averaged over the six different wind turbine columns. \label{fig:flowfield_statistics}}
\end{figure}

@@

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals.

@@

The figure shows that the first row (R1) holds by far the most promise for optimizing wind-farm power. This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}). At the other end of the spectrum, the last row (R12) is the least useful. The intermediate rows (R2--R11) lie closer together, with the general trend being that the potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous.

@@

In the first case, the controls, which were specifically generated for selected turbines, are reassigned to other turbines by randomly swapping the control sets of different turbine columns. In doing so, each turbine will receive controls that were specifically designed for another turbine in the same row. To avoid erroneous conclusions based on coincidence, the column swap is performed in 2 random independent ways. Row-averaged power for these cases is shown in Fig.~\ref{fig:scrambled}a.

In the second case, controls remain assigned to their original turbines, but are shuffled in time by randomly swapping optimal controls generated for different control windows. In this way, the spectral thrust characteristics for timescales smaller that the control horizon $T_A = 120$ s remain unchanged, whereas the time synchronization of control actions to specific flow events is eliminated. Similar to the first case, this is done in 2 random ways, and Fig.~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be attained by an NREL 5MW rotor with a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A). Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figure14}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

\subsubsection{Full-scale wind-farm test}
In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm, corresponding to the full $12 \times 6$ aligned wind farm of MM17. Simulations are performed for a reduced range of amplitudes and Strouhal numbers, corresponding to the most favorable region identified in the parameter sweeps above. In order to increase statistical convergence, the time horizon for each simulation is extended to a physical time of 10 hours.

Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale test. Figure~\ref{fig:sinus_fullscale}a shows the relative power gains over the reference case for the full wind farm. It can be seen that the total power gain or loss is below 0.5\% for each of the sinusoidal control cases. Figure~\ref{fig:sinus_fullscale}b shows the row-wise power extraction for the reference case, the sinusoidal thrust case with $(St, A) = (0.25, 1.5)$, and the optimal control case C3t5. It is shown that, although the second row of the sinusoidal thrust case achieves similar power gains as those observed above, from the fifth row onwards power is slightly reduced in the sinusoidal case.

@@

The top panel of Fig.~\ref{fig:cross_section_sinus} shows cross sections of time-averaged axial velocities $\widetilde{U}_x$ at the rotor disk locations for the reference case. Further, the middle and bottom panel illustrate deviations from the reference velocity for the $(St^\star, A^\star)$ sinusoidal case and the optimal control case C3t5 respectively.  The figure shows that both controlled cases show similar characteristics at the second turbine row, with an increased axial velocity at the rotor disk, accompanied by decreased velocities above and below. Downstream, it can be seen that the passive turbines of the sinusoidal case fail to retain increased velocities at the rotor disks, instead resulting in slightly lower disk velocities starting from the fifth row. In contrast, case C3t5, in which all turbines are actively controlled, succeeds to attain similar cross section characteristics with higher rotor velocities in the downstream as well.

@@

As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row. However, important comments should be made. First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading. Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach. Second, even though experiments have shown that, for practically relevant tip speed ratios, wind turbines shed vortices in a similar way as disk-like bluff bodies \citep{medici2006measurements}, the current behavior could still be an artifact of the relatively simple ADM used throughout this study. Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.

>>>>

\newcommand{\revision}[1]{{\color{red} #1}}

@@

\subsection{Case setup}
The wind farm considered in MM17 has an aligned pattern of 12 rows by 6 columns. \revision{The wind turbines have a hub height $z_h = 100$~m with a rotor diameter $D = 100$~m, and are spaced apart by 6$D$ in both axial and transversal directions.} The flow is simulated on a domain of $10 \times 3.6 \times 1$ km$^3$, discretized on a simulation grid of $384 \times 192 \times 144$ grid points. A snapshot of the streamwise velocity field is shown in Fig. \ref{fig:flowfield_MM17}. The wind farm was controlled over a total of $N_A = 15$ time windows with a prediction horizon $T = 240$ s (\revision{i.e. the time it takes for the flow to pass four rows of turbines}) and a flow advancement time of $T_A = T/2 = 120$ s, resulting in a total control time $T_{\rm tot} = N_A T_A = 30$ minutes.

@@

A conventionally (greedily) controlled wind farm with steady $C_T' = 2$ (corresponding to the Betz limit in momentum theory) was defined as a reference case. Several different optimal control cases were defined, based on the wind-turbine response time $\tau = $ 0, 5, or 30 s (instantaneous, fast, or slow response, see Eq. \ref{eq:ctihat_constraint}) and the maximal thrust coefficient $C_{T,\text{max}}' = 2$ or~3 \revision{, with thrust forces that can respectively only be reduced (underinductive), or also increased (overinductive) compared to the Betz optimum at $C_T' = 2$} (see Eq. \ref{eq:boxct_constraint}). Cases are denoted as C<X>t<Y>, where X and Y represent $C_{T,\text{max}}'$ and $\tau$ respectively, e.g. C3t30  for the case with $C_{T,\text{max}}' = 3$ and $\tau = 30$ s. \revision{The choice of (and sensitivity to) setup parameters is further elaborated in MM17.}

@@

Figure \ref{fig:flowfield_statistics}c shows sideviews of turbulence kinetic energy $k$. The figure shows an increase in turbulence throughout the entire wind farm, spreading to the internal boundary layer above the turbines. Note specifically the sharp increase in turbulence in the core wake region behind the first-row turbine, for which an enhanced recovery was found as discussed above. \revision{The turbulence intensity $TI \equiv (2k/3  )^{1/2}/(U_x^2+U_y^2+U_z^2)^{1/2}$ at hub height (not shown in the figure) is 10\% at the inlet for both the reference case and the controlled case. The combination of reduced near-wake mean velocities and increased velocity fluctuations in the controlled case increase local $TI$ in the turbine wakes (ranging from $\approx 2$\%-points in the wakes of middle rows to $\approx 12$\%-points in the first and last rows). This increase in turbulence intensity dissipates to below $1$\%-point difference at 10$D$ downstream of the last row.}

@@

\begin{figure}
	\includegraphics[width=\textwidth]{figure4}
	\caption{Time-averaged flow field quantities of simulation results from MM17. \emph{Left (1):} Time averages of reference simulations with $C_T' = 2$ for all turbines. \emph{Right (2):} Difference $\Delta$ between reference and optimal control (C3t5) simulation, defined as $\Delta = X_{\rm C3t5} - X_{\rm REF}$ for any variable $X$. \emph{a):} Axial velocity $U_x$ (planview at hub height). \emph{b):} Axial velocity $U_x$ (sideview through turbine column). \emph{c):} Turbulence kinetic energy $k = (1/2)(\overline{\widetilde{u}'_x \widetilde{u}'_x} + \overline{\widetilde{u}'_y \widetilde{u}'_y} + \overline{\widetilde{u}'_z \widetilde{u}'_z})$ (sideview). \emph{d):} Turbulence top-down transport $ - \overline{\widetilde{u}'_x\widetilde{u}'_z}$. \emph{e):} Mean-flow top-down transport $ - \widetilde{U}_x \widetilde{U}_z$ (sideview). \emph{f):} Turbulence horizontal transport $\overline{\widetilde{u}'_x \widetilde{u}'_y}$ (planview). \emph{g):} Mean-flow horizontal transport $\widetilde{U}_x \widetilde{U}_y$ (planview). Black lines indicate wind turbine locations. Simulation results are averaged over the six different wind turbine columns. \label{fig:flowfield_statistics}}
\end{figure}

@@

The current section focuses on the optimal thrust coefficients generated by the optimal control simulations in MM17, and performs numerical experiments to uncover some of the characteristics of these control signals. \revision{Note that the conclusions drawn within this section should be interpreted as observations of the current C3t5 optimal control cases, given specific wind-farm layout and flow conditions, and hence cannot just be generalized for any wind-farm control in general.}

@@

The figure shows that the first row (R1) holds by far the most promise for optimizing wind-farm power. This is not surprising as R1 produces the most power of all rows, and typically leaves the deepest wakes, causing second-row turbines to perform  poorly in aligned wind-farm layouts (see, e.g., \citealp{porte2013numerical,nilsson2015large}; and \citealp{stevens2016effects}). At the other end of the spectrum, the last row (R12) is the least useful. The intermediate rows (R2--R11) lie closer together, with the general trend being that the potential is somewhat decreased with downstream distance into the wind farm, although this decrease is not monotonous. \revision{Note that these observations are in contrast to the study of \cite{ciri2017large}, where extremum seeking control is applied to optimize individual turbines in an array of three turbines. They found that set points in the first-row turbines remain untouched from their locally optimal values whereas downstream turbines are self-optimized for wake conditions. This contrast can be attributed to our dynamic and simultaneous wind-farm control approach, which differs from their sequential individual set-point control. Furthermore, self-optimization in the latter study is achieved by recalibrating the torque gain. The baseline controller in our current actuator disk model makes abstraction of pitch and generator torque control and keeps $C_T'$ at its theoretical optimum instead. }

@@

In the first case, the controls, which were specifically generated for selected turbines, are reassigned to other turbines by randomly swapping the control sets of different turbine columns. In doing so, each turbine will receive controls that were specifically designed for another turbine in the same row. To avoid erroneous conclusions based on coincidence, the column swap is performed in 2 random independent ways. \revision{The variability of flow conditions for different columns can be qualitatively observed in Figure 2. To further strengthen the hypothesis of the current experiment, we verified that the  correlation between flow conditions in different columns is small, i.e. with an average Pearson correlation coefficient of 0.12 between columns for the incoming velocity fluctuations 6$D$ upstream of the first row.} Row-averaged power for these cases is shown in Fig.~\ref{fig:scrambled}a.

In the second case, controls remain assigned to their original turbines, but are shuffled in time by randomly swapping optimal controls generated for different control windows. In this way, the spectral thrust characteristics for timescales smaller that the control horizon $T_A = 120$ s remain unchanged, whereas the time synchronization of control actions to specific flow events is eliminated. Similar to the first case, this is done in 2 random ways, \revision{and the limited correlation between velocity fluctuations in different time windows was quantified at 0.07}. Figure ~\ref{fig:scrambled}b illustrates the row-averaged power for these cases.

@@

Figure~\ref{fig:sinus_baseline} illustrates the power extraction for all cases considered. Figure~\ref{fig:sinus_baseline}a illustrates the relative power gains over the reference case. From the figure it can be seen that there is a well-defined range of values for $A$ and $St$ for which wind-farm power can be increased substantially through upstream sinusoidal thrust variations, with a maximal power increase of $\approx$ 5\% at $(St^\star, A^\star) = (0.25, 1.5)$. For instance, a Strouhal number $St = 0.25$ corresponds here to a sine wave period of $\approx$ 50 s for a turbine with diameter $D = 100$ m and a freestream velocity $U_\infty = 8.5$ m s$^{-1}$. The resulting maximum thrust coefficient of 3.5 can, e.g., be \revision{attained by the NREL 5MW turbine with a slight modification in the rotor design}, i.e. a 50\% increase in chord length at a tip speed ratio that is 25\% higher than the greedy tip speed ratio (see \citealp{goit2015optimal}, Appendix A). \revision{Furthermore, given this redesign, dynamic reductions from this maximum value could be realized through pitch control. More specifically, based on speculated specifications of similar turbines in practice, the NREL 5 MW is designed with a pitch rate of $8^{\circ}/$s. This implies that the rotor can go from power maximization ($0^\circ$ pitch) to total curtailment in fully feathered blade position ($90^\circ$ pitch) in 11.25 seconds \cite{jonkman2009definition}.} Figure~\ref{fig:sinus_baseline}b illustrates normalized power extraction by row for the reference case, the best sinusoidal case, and the first four rows of the optimal control case C3t5 from MM17. The figure shows that the power gain in the sinusoidal cases originates mostly from the second row, and that power in the first row is decreased by approximately 5\%. In contrast, optimal control case C3t5, in which all rows are active, also increases power in rows 3 and 4, and reduces first-row power by only 1\%. 
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figure14}
	\caption{Power extraction of baseline sinusoidal thrust case ($S = 6D$, $z_0 = 10^{-1}$ m). \emph{a) } Relative gain in mean wind-farm power extraction over reference case as a function of sine amplitude $A$ and frequency $St$. \emph{b) } Row-averaged mean power extraction for the best sinusoidal case, the optimal control case C3t5 from MM17, and the reference case, normalized by first-row reference power.\label{fig:sinus_baseline} }
\end{figure}

@@

\subsubsection{\revision{Full-scale wind-farm LES}}
In the remainder of this section, the sinusoidal variation strategy will be tested in a full-scale wind farm \revision{LES}, corresponding to the full $12 \times 6$ aligned wind farm of MM17. Simulations are performed for a reduced range of amplitudes and Strouhal numbers, corresponding to the most favorable region identified in the parameter sweeps above. In order to increase statistical convergence, the time horizon for each simulation is extended to a physical time of 10 hours.

Figure~\ref{fig:sinus_fullscale} shows the power extraction of the full-scale \revision{LES}. Figure~\ref{fig:sinus_fullscale}a shows the relative power gains over the reference case for the full wind farm. It can be seen that the total power gain or loss is below 0.5\% for each of the sinusoidal control cases. Figure~\ref{fig:sinus_fullscale}b shows the row-wise power extraction for the reference case, the sinusoidal thrust case with $(St, A) = (0.25, 1.5)$, and the optimal control case C3t5. It is shown that, although the second row of the sinusoidal thrust case achieves similar power gains as those observed above, from the fifth row onwards power is slightly reduced in the sinusoidal case.

@@

The top panel of Fig.~\ref{fig:cross_section_sinus} shows cross sections of time-averaged axial velocities $\widetilde{U}_x$ at the rotor disk locations for the reference case. Further, the middle and bottom panel illustrate deviations from the reference velocity for the $(St^\star, A^\star)$ sinusoidal case and the optimal control case C3t5 respectively.  The figure shows that both controlled cases show similar characteristics at the second turbine row, with an increased axial velocity at the rotor disk, accompanied by decreased velocities above and below. Downstream, it can be seen that the passive turbines of the sinusoidal case fail to retain increased velocities at the rotor disks, instead resulting in slightly lower disk velocities starting from the fifth row. In contrast, case C3t5, in which all turbines are actively controlled, succeeds to attain similar cross section characteristics with higher rotor velocities in the downstream as well. \revision{Note also that, for the fifth row, the disk velocity is slightly lower for the sinusoidal control case than for the reference case, consistent with the decreased power extraction observed in Fig. \ref{fig:sinus_fullscale}. This can be explained by the fact that first-row control actions cause enhanced entrainment of momentum from the internal boundary layer above the turbine canopy that would otherwise be entrained by natural turbulent mixing in passive downstream rows. In consequence, lesser entrainment occurs for downstream rows, resulting in a slight decrease in disk velocities from the fifth row onwards.}

@@

As shown throughout the current section, a qualitative analysis of instantaneous flow features in the optimal induction control wind farm from MM17 has led to the identification of a sinusoidal thrust control strategy for first-row turbines, resulting in increased power extraction in the second row. However, important comments should be made. First, sustained sinusoidal thrust variations with a large amplitude could contribute significantly to turbine fatigue loading \revision{of the first-row turbines. Furthermore, partial wake alleviation and unsteady passing of abovementioned vortex rings could also increase fatigue loading in downstream rows.}. Hence, structural aspects should be taken into account upon evaluating the practical viability of the approach. Second, even though experiments have shown that, for practically relevant tip speed ratios, wind turbines shed vortices in a similar way as disk-like bluff bodies \citep{medici2006measurements}, the current behavior could still be an artifact of the relatively simple ADM used throughout this study. Further verification using higher fidelity wind turbine models, such as actuator line models, and wind-tunnel testing is hence necessary.
